---
title: 'Reproducible Research: Peer Assessment 2'
author: "Kevin Tham"
date: "April 12, 2018"
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

## Synopsis

We address the following questions through analysis of the NOAA Storm Database:

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?



## Data Processing

First, we load the required packages and install them first if they are not available:

```{r}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(knitr, dplyr, ggplot2, tidyr, hexbin, timeDate)
```

Next we download and unzip the data file to read in the dataset:

```{r, cache=TRUE}
destfile <- 'repdata%2Fdata%2FStormData.csv.bz2'
url <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'

if (!file.exists(destfile))
  download.file(url, destfile, method = "auto")

df <- read.csv(destfile)
```

We will explore the dataset using the `head()` and `str()` functions:

```{r}
head(df)
str(df)
```

There are many variables present in this dataset, not all of which will be relevant to our analysis. We recap our goal:

1. To find out which **EVTYPE**(-s) is most harmful to **population health** across the US
2. To find out which **EVTYPE**(-s) have the greatest **economic consequences**

Therefore, we will narrow down the number of variables to those that are related to **EVTYPE**, **population health** and **economic consequences**.

Population health: 

* `FATALITIES`
* `INJURIES` 

Economic consequences: 

* `PROPDMG` 
* `PROPDMGEXP` 
* `CROPDMG` 
* `CROPDMGEXP` 

We will select these variables from the original dataset in addition to `EVTYPE` for obvious reasons, and `STATE__` in order to analyse the spatial variation of population health and economic consequences:

```{r}
df_filtered <- select(df, EVTYPE, STATE__, FATALITIES, INJURIES, 
                      PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)
```

The proportion of missing values in the dataframe is first determined for the relevant variables:

```{r}
mean(is.na(df_filtered))
```

Surprisingly there is a complete absence of missing variables.

Next we examine the variable `EVTYPE`:

```{r, results='hide'}
unique(df_filtered$EVTYPE)
```

It is clear that there are many duplicates, errors and variants present in the variable `EVTYPE`. Therefore we will proceed by cleaning up this variable, using regular expressions to substitute variable names with the correct variable.

```{r}
event <- df_filtered$EVTYPE
event <- gsub('.*TSTM.*', 'THUNDERSTORM WIND', event, ignore.case=TRUE)
event <- gsub('.*THUNDER.*', 'THUNDERSTORM WIND', event, ignore.case=TRUE)
event <- gsub('.*THUND.*', 'THUNDERSTORM WIND', event, ignore.case=TRUE)
event <- gsub('.*micro.*', 'THUNDERSTORM WIND', event, ignore.case=TRUE)
event <- gsub('.*SPOUT.*', 'WATERSPOUT', event, ignore.case=TRUE)
event <- gsub('.*TORNADO.*', 'TORNADO', event, ignore.case=TRUE)
event <- gsub('^HIGH WIND.*', 'HIGH WIND', event, ignore.case=TRUE)
event <- gsub('.*AVALANCHE.*', 'AVALANCHE', event, ignore.case=TRUE)
event <- gsub('.*BLIZZARD.*', 'BLIZZARD', event, ignore.case=TRUE)
event <- gsub('.*SNOW.*', 'HEAVY SNOW', event, ignore.case=TRUE)
event <- gsub('.*HURRICANE.*', 'HURRICANE/TYPHOON', event, ignore.case=TRUE)
event <- gsub('.*TYPHOON.*', 'HURRICANE/TYPHOON', event, ignore.case=TRUE)
event <- gsub('.*HEAT.*', 'HEAT', event, ignore.case=TRUE)
event <- gsub('.*WARM.*', 'HEAT', event, ignore.case=TRUE)
event <- gsub('.*HIGH TEMP.*', 'HEAT', event, ignore.case=TRUE)
event <- gsub('.*COLD.*', 'EXTREME COLD/WIND CHILL', event, ignore.case=TRUE)
event <- gsub('.*HAIL.*', 'HAIL', event, ignore.case=TRUE)
event <- gsub('.*flash.flood.*', 'FLASH FLOOD', event, ignore.case=TRUE)
event <- gsub('.*flash.*', 'FLASH FLOOD', event, ignore.case=TRUE)
event <- gsub('.*LIGHTNING.*', 'LIGHTNING', event, ignore.case=TRUE)
event <- gsub('(?<!FLASH )FLOOD.*', 'FLOOD', event, ignore.case=TRUE, perl=TRUE)
event <- gsub('.*FLOOD.*', 'FLOOD', event, ignore.case=TRUE)
event <- gsub('.*FUNNEL.*', 'FUNNEL CLOUD', event, ignore.case=TRUE)
event <- gsub('.*FIRE.*', 'WILDFIRE', event, ignore.case=TRUE)
event <- gsub('.*TROPICAL STORM.*', 'TROPICAL STORM', event, ignore.case=TRUE)
event <- gsub('.*CURRENT.*', 'RIP CURRENT', event, ignore.case=TRUE)
event <- gsub('.*DRY.*', 'DROUGHT', event, ignore.case=TRUE)
event <- gsub('(?<!FREEZING )RAIN.*', 'HEAVY RAIN', event, ignore.case=TRUE, perl=TRUE)
event <- gsub('.*RAIN.*', 'HEAVY RAIN', event, ignore.case=TRUE)

table(grep('(?<!freezing )rain',event,ignore.case=TRUE,value=TRUE, perl=TRUE))
unique(event)
```

## Results

## Conclusion
